---
title: "Distribution script.qmd"
format: html
editor: visual
---

## Introduction

Script that explores the distribution of both African penguins and caracals and visualises whether they overlap in their sighted regions. The data is found from INaturalist and based in the bounds of the Western Cape.

## Running Code

```{r}#libraries}
library(rinat)#retrieve the data from INAT
library(sf)##make spatial object
library(leafpop)
##add a baselayer to the plot
library(rosm)
library(ggspatial)
library(prettymapr)
##generates interactive maps
library(leaflet)
library(htmltools)
##common sense check 
library(mapview)##handles layers and legends
library(leafpop)

```

**Data wrangling**

```{r}
#Call the data directly from iNat
peng <- get_inat_obs(taxon_name = "African penguin",
                   bounds = c(-35.2, 16.3, -30.5, 24),#bounds of the Western Cape
                   maxresults = 1000)
head(peng)
names(peng)

```

```{r}
##filter to get quality grade research
peng <- peng %>% filter(positional_accuracy<46 & 
                      latitude<0 &
                      !is.na(latitude) &
                      captive_cultivated == "false" &
                      quality_grade == "research")
```

```{r}
##double check the data frames crs
st_crs(peng)##4326

#make the penguin data frame a spacial object of class sf
peng <- st_as_sf(peng, coords = c("longitude", "latitude"), crs = 4326)
```

```{r}
#plot the basic points inside the bounds
ggplot() + geom_sf(data=peng)
```

```{r}
##add a basemap to see more of the coastline and features
ggplot() + 
  annotation_map_tile(type = "thunderforestoutdoors", progress = "none") + 
  geom_sf(data=peng)

rosm::osm.types()##tyes of baselayer maps

```

```{r}
#generate an interactive map
leaflet() %>%
  # Add default OpenStreetMap map tiles
  addTiles(group = "Default") %>%  
  # Add our points
  addCircleMarkers(data = peng,
                   group = "African Penguin",
                   radius = 3, 
                   color = "purple") 
```

##do a common sense check on the data to make sure that it is legitemate sightings

##cause it was taken from INat

```{r}
mapview(peng, 
        popup = 
          popupTable(peng,
                     zcol = c("user_login", "captive_cultivated", "url")))
```

```{r}
lpeng <- peng %>%
  mutate(click_url = paste("<b><a href='", url, "'>Link to iNat observation</a></b>"))

```

```{r}
mapview(peng, 
        popup = 
          popupTable(lpeng,
                     zcol = c("user_login", "captive_cultivated", "click_url")))
```

**Overlay caracal sightings data**

```{r}
##draw caracal data from INat
cara <- get_inat_obs(taxon_name = "Caracal",
                     bounds = c(-35.2, 16.3, -30.5, 24),
                     maxresults = 1000)
```

```{r}
#filter, shape and sort out caracal data from INat
cara <- cara %>% filter(positional_accuracy<46 & 
                          latitude<0 &
                          !is.na(latitude) &
                          captive_cultivated == "false" &
                          quality_grade == "research")
```

```{r}
##double check the data frames crs
st_crs(cara)##4326 same as peng

#make the penguin data frame a spacial object of class sf
cara <- st_as_sf(cara, coords = c("longitude", "latitude"), crs = 4326)
```

```{r}
peng$type <- "African Penguins"
cara$type <- "Caracals"
```

```{r}
pc <- rbind(peng, cara)

ggplot(combined) +
  geom_sf(aes(color = type), alpha = 0.7) +
  theme_minimal()## created an overlayed plot
```

```{r}
leaflet() %>%
  # Add default OpenStreetMap map tiles
  addTiles(group = "Default") %>%  
  # Add our points
  addCircleMarkers(data = peng,
                   group = "African Penguin",
                   radius = 3, 
                   color = "purple") %>%
  # Cara points
  addCircleMarkers(data = cara,
                   group = "Cara",
                   radius = 3,
                   color = "orange",
                   fillOpacity = 0.7) %>%
  addLayersControl(
    overlayGroups = c("African Penguin", "Cara"),
    options = layersControlOptions(collapsed = FALSE)
  )  ##this adds a layer so that you can toggle the two species on and off

##interactive map of two species
```

**make interactive map with clickable URL groups**

```{r}
#Load South Africa land polygons 
sa_land <- ne_countries(scale = "medium", country = "South Africa", returnclass = "sf")

```

```{r}
##Convert your data to sf points again to ensure they are in the same crs 
peng <- st_as_sf(peng, coords = c("lon", "lat"), crs = 4326)
cara <- st_as_sf(cara, coords = c("lon", "lat"), crs = 4326)


```

```{r}
#Filter to keep only the caracal pts recorded on land as INat sourced data has a few ocean sightings which would make no sense
cara <- cara[st_intersects(cara, sa_land, sparse = FALSE), ]

```

```{r}
##Add clickable URL popups for both peng and cara
peng <- peng %>%
  mutate(click_url = paste0("<b><a href='", url, "'>Link to iNat observation</a></b>"))

```

```{r}
cara <- cara %>%
  mutate(click_url = paste0("<b><a href='", url, "'>Link to iNat observation</a></b>"))
mp_peng <- mapview(peng,
                   col.regions = "lightgreen",
                   popup = popupTable(peng, zcol = c("user_login", "captive_cultivated", "click_url")),
                   layer.name = "African Penguins")

```

```{r}
mp_cara <- mapview(cara,
                   col.regions = "orange",
                   popup = popupTable(cara, zcol = c("user_login", "captive_cultivated", "click_url")),
                   layer.name = "Caracals")

```

```{r}
##Combine the layers
mp_peng + mp_cara##Final map with legend and open street map view(select)
```
